---
title: "color patch"
author: "Andrew O. Rubio"
date: "11/11/2022"
output: html_document
---

```{r}
library(tximport)
library(DESeq2)
library(dplyr)
library(foreach)
library(data.table)
library(splines)
library(ggthemes)
library(scales)
library(gridExtra)
library(tidyr)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library("BiocParallel")
register(SnowParam(8))
library(apeglm)
library(tidyverse)
library(topGO)
library(GO.db)
library(tidyverse)
library(SummarizedExperiment)
```

Some functions for this analysis.
```{r}
# this function pulls out particular a priori genes of interest into a new data frame
aprioris <- function(df, annos){
  df.aprioris <- data.frame()
  for (i in 1:length(colors$gene_name)){
    searchterm <- paste0("\\b", colors$gene_name[i], "\\b")
    searchterm <- tolower(searchterm)
    tmp <- df %>% filter(str_detect(Gene, searchterm))
    df.aprioris <- rbind(df.aprioris, tmp)
}
   return(df.aprioris)
}

# this function extracts all genes below a specified alpha value, prints the number of DE genes to screen, and saves a spreadsheet to file
siggies <- function(df, alpha, csv){
  sigs <- df %>% filter(padj < alpha)
  print(paste0("Number of significant genes: ", length(sigs$padj)))
  write.csv(sigs, csv, row.names = FALSE)
  return(sigs)
}


# significant gene wrapper function
SigGeneWrapper <- function(model_output, alpha, comparison){
  print(paste0("Running ", comparison, " with alpha = ", alpha))
  df <- setDT(as.data.frame(model_output), keep.rownames = "Gene")
  # add annotation data
  #df1 <- dplyr::left_join(df, annos, by = "transcript") 
  # get significant genes
  print("Overall significant genes")
  sigs <- siggies(df, alpha, paste0("results/skin/", comparison, "_genes.csv"))
  # add annotation data
  colordf <- aprioris(df, annos)
  #get significant color genes
  print("Significant color genes")
  color.sigs <- siggies(colordf, alpha, paste0("results/skin/", comparison, "_colorgenes.csv"))
}

```

#Data import

```{r}
#sessionInfo()
```

```{r}
###### list all samples from expression data ####
# get the directory/path for each sample in this study
base_dir <- getwd()
filenames <- list.files(path = "skin_counts/", full.names = F, recursive = F)
files <- file.path(base_dir, "skin_counts/", filenames) # files = directory + salmon directory + sample name + quantifictaion file name
names(files) <- "~/Google Drive/Projects/Dissertation/fungi vs carotenoids/skin_counts/"
all(file.exists(files)) # do these all actually exist?
list(files)

#### make sample spreadsheet for black vs color####
samples_color <- as.data.frame(filenames)

# get sample

samples_color$sample <- filenames %>% gsub(pattern = "_", replacement = "") %>% gsub(pattern = ".gene.counts", replacement = "") %>% gsub(pattern = "black", replacement = "") %>% gsub(pattern = "One", replacement = "") %>% gsub(pattern = "varadero", replacement = "")%>% gsub(pattern = "Yeast", replacement = "") %>% gsub(pattern = "sauce", replacement = "") %>% gsub(pattern = "red", replacement = "") %>% gsub(pattern = "orange", replacement = "") %>% gsub(pattern = "C", replacement = "") %>% gsub(pattern = "liver", replacement = "") %>% gsub(pattern = "liver", replacement = "") %>% gsub(pattern = "L", replacement = "") %>% gsub(pattern = "Y", replacement = "") %>% gsub(pattern = "B", replacement = "")

samples_color$sample

# get morph
samples_color$morph <- filenames %>% gsub(pattern = "_", replacement = "") %>% gsub(pattern = ".gene.counts", replacement = "") %>% gsub(pattern = "Yeast", replacement = "") %>% gsub(pattern = "black", replacement = "") %>% gsub(pattern = "orange", replacement = "") %>% gsub(pattern = "[0-9]", replacement = "") %>% gsub(pattern = "red", replacement = "") %>% gsub(pattern = "C", replacement = "") %>% gsub(pattern = "", replacement = "") %>% gsub(pattern = "L", replacement = "") %>% gsub(pattern = "Y", replacement = "") %>% gsub(pattern = "C", replacement = "") %>% gsub(pattern = "B", replacement = "") %>% gsub(pattern = "liver", replacement = "")

samples_color$morph

# get tissue

samples_color$tissue <- filenames %>% gsub(pattern = "_", replacement = "") %>% gsub(pattern = ".gene.counts", replacement = "") %>% gsub(pattern = "C", replacement = "") %>% gsub(pattern = "Yeast", replacement = "") %>% gsub(pattern = "[0-9]", replacement = "") %>%  gsub(pattern = "Y", replacement = "") %>% gsub(pattern = "B", replacement = "Black") %>%  gsub(pattern = "L", replacement = "Liver") %>% gsub(pattern = "sauceyellow", replacement = "orange") %>% gsub(pattern = "varaderoyellow", replacement = "red") %>% gsub(pattern = "sauce", replacement = "") %>% gsub(pattern = "varadero", replacement = "") %>% gsub(pattern = "black", replacement = "")  %>% gsub(pattern = "yellow", replacement = "")  %>% gsub(pattern = "liver", replacement = "") 

samples_color$tissue 

# get diet

samples_color$diet <- filenames %>% gsub(pattern = "_", replacement = "") %>% gsub(pattern = ".gene.counts", replacement = "") %>% gsub(pattern = "[0-9]", replacement = "") %>% gsub(pattern = "C", replacement = "carotenoid") %>% gsub(pattern = "sauce", replacement = "") %>% gsub(pattern = "varadero", replacement = "") %>% gsub(pattern = "orange", replacement = "") %>% gsub(pattern = "liver", replacement = "") %>% gsub(pattern = "black", replacement = "") %>% gsub(pattern = "yellow", replacement = "") %>% gsub(pattern = "L", replacement = "") %>% gsub(pattern = "B", replacement = "") %>% gsub(pattern = "Yeast", replacement = "yeast") %>% gsub(pattern = "Y", replacement = "")

samples_color$diet

#samples_color$pair.n <- as.factor(rep(c(1:3), 2))

samples_color$pair.nn <- c(("1"),("1"),("2"),("2"),("3"),("3"),("1"),("1"),("2"),("2"),("3"),("3"))
samples_color$diet.color <- c(("cb"),("co"),("cb"),("co"),("cb"),("co"),("cb"),("co"),("cb"),("co"),("cb"),("co"),("yb"),("yo"),("yb"),("yo"),("yb"),("yo"),("yb"),("yo"),("yb"),("yo"),("yb"),("yo"))



deseqsamples_color <- samples_color [,-1]
deseqsamples_color <- deseqsamples_color [,-1]
list(deseqsamples_color)
write.table(deseqsamples_color, "r-output/.tsv", row.names = TRUE, sep = "\t")
```

# Import expression data from.
```{r}
annos <- fread("R_imi.UNK.annotations.genesymbol.tsv", header = FALSE)
colnames(annos) <- c("transcript", "Gene")
annos$Gene <- tolower(annos$Gene)

# rname columns from annos file
# import data
countdata <- data.frame(annos$transcript)
colnames(countdata) <- "transcript"
for (counts in (1:length(filenames))) {
  # import data for that sample
  toimport <- filenames[counts]
tmpdata <- read.table(paste0("skin_counts/", toimport), header = FALSE, sep = "\t")
  # sample id
  samplename <- toimport %>% gsub(pattern = ".gene.counts", replacement = "")
  colnames(tmpdata) <- c("transcript", samplename)
  countdata <- dplyr::left_join(countdata, tmpdata, by = "transcript")
}


dim(countdata)

write.csv(countdata, "r-output/skin_count_data.csv", row.names = F)


countdata <- data.frame(countdata, row.names = 1)
head(countdata)
head(annos)
```

# Import a priori candidate color gene
```{r}
colors <- read.csv("master_gene_list_2024_with_added_genes.csv", header = TRUE)
colnames(colors) <- "gene_name" 
colors$gene_name <- tolower(colors$gene_name)
head(colors)
```

# Quick data control. Change and "NA" values to "0". Also remove any row with total expression <50 (or 1 count/every 3rd sample).
```{r}
# change all NA to 0
print(paste("Number of NA after import:", sum(is.na(countdata))))
countdata[is.na(countdata)] <-  0
print(paste("Number of NA after NA removal:", sum(is.na(countdata))))

# remove any row with total expression < 10
print(paste("Number of rows (transcripts) after import:", nrow(countdata)))
keep <- rowSums(countdata) >= 50
countdata <- countdata[keep,]
print(paste("Number of rows (transcripts) after filter:", nrow(countdata)))

head(countdata)
countdata
write.csv(countdata, "skin_name_count_data.csv", row.names = F)
```

```{r}
# make rownames column 1
countdata <- setDT(countdata, keep.rownames = "transcript")[]

# add gene symbol
countdata2 <- dplyr::left_join(annos, countdata, by = "transcript")

# change all NA to 0
print(paste("Number of NA after import:", sum(is.na(countdata2))))
countdata2[is.na(countdata2)] <-  0
print(paste("Number of NA after NA removal:", sum(is.na(countdata2))))


# combine all counts that map to the same gene
countdata3 <- aggregate(countdata2[, 3:26], list(countdata2$Gene), sum)
countdata3[is.na(countdata3)] <-  0
colnames(countdata3)[1] <- "Gene"
countdata <- countdata3

countdata <- data.frame(countdata, row.names = 1)

# save gene-level count data
write.table(countdata, "r-output/gene_name_skin_count.data.tsv", row.names = TRUE, sep = "\t")
```

-----------
Sauce morph: all carotenoid skin vs all yeast skin
-----------
```{r}
dds_all_sauce <- DESeqDataSetFromMatrix (countData = countdata,
                                       colData = deseqsamples_color,
                                       design = ~ morph + morph:pair.nn + morph:diet)
#run Wald test
dds.sauce.yellowVSblack <- DESeq(dds_all_sauce)
resultsNames(dds.sauce.yellowVSblack)

#extract the DEG between orange and black skin patches in the Sauce morph
res.sauce.yellowVSblack <- results(dds.sauce.yellowVSblack, name = c("morphsauce.dietyeast"))
res.sauce.yellowVSblack$transcript <- mcols(res.sauce.yellowVSblack)$transcript

# how many are "significant"?
table(res.sauce.yellowVSblack[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(res.sauce.yellowVSblack, 0.05, "sauce.carotenoidvsyeast")

#sauce heatmaps:melanin, pteridine, and carotenoid

sauce.skincount <- dplyr::select(countdata, contains("sauce"))
sauce.skinmorph <- deseqsamples_color %>% filter(morph == "sauce")

sauce.skin <- DESeqDataSetFromMatrix (countData = sauce.skincount,
                                       colData = sauce.skinmorph,
                                       design = ~ diet)

sauce.significant.color.vsd <- varianceStabilizingTransformation(sauce.skin, blind=FALSE)
sauce.significant.color.mat<- assay(sauce.significant.color.vsd)
pheatmap(sauce.significant.color.mat)
         
#melanin
sauce.skin<- c("phgdh", "pdgfb", "creb3l1","sult1a1", "lpl", "atp6v1e1", "dhrs3", "gfpt1", "hdac1")

pheatmap(sauce.significant.color.mat[sauce.skin,], cluster_cols = F, cluster_rows = T, scale = "row",labels_col = c("Carotenoid","Yeast"),main = "", border_color="Black", color = colorRampPalette(c("grey 60", "white", "orange"))(50), angle_col = "0",annotation_legend = F, gaps_col = c(),fontsize= 25, filename = "Figures/sauce.skin.carotenoidvsyeast.png")
```

--------------
Varadero morph: all carotenoid skin vs all yeast skin
--------------
```{r}
dds_all_varadero <- DESeqDataSetFromMatrix (countData = countdata,
                                       colData = deseqsamples_color,
                                       design = ~ morph + morph:pair.nn + morph:diet)

#run Wald test
dds.varadero.redVSblack <- DESeq(dds_all_varadero, test="Wald")
resultsNames(dds.varadero.redVSblack)

#extract the DEG between black and red skin patches in the Varadero morph
res.varadero.redVSblack <- results(dds.varadero.redVSblack, name = c("morphvaradero.dietyeast"))
res.varadero.redVSblack$transcript <- mcols(res.varadero.redVSblack)$transcript
# how many are "significant"?
table(res.varadero.redVSblack[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(res.varadero.redVSblack, 0.05, "varadero.carotenoidvsyeast")

#Varadero heatmaps:melanin, pteridine, and carotenoid
varadero.skincount <- dplyr::select(countdata, contains("Varadero"))
varadero.skinmorph <- deseqsamples_color %>% filter(morph == "varadero")

varadero.significant.color.dds <- DESeqDataSetFromMatrix (countData = varadero.skincount,
                                       colData = varadero.skinmorph, 
                                       design = ~ diet)

varadero.significant.color.vsd <- varianceStabilizingTransformation(varadero.significant.color.dds, blind=FALSE)
varadero.significant.color.mat<- assay(varadero.significant.color.vsd)


#melanin
varadero.skin<- c("sult1a1", "arcn1")

pheatmap(varadero.significant.color.mat[varadero.skin,], cluster_cols = F, cluster_rows = T, scale = "row",labels_col = c("Carotenoid","Yeast"),main = "", border_color="Black", color = colorRampPalette(c("grey 60", "white", "red"))(50), angle_col = "0",annotation_legend = F, gaps_col = c(),fontsize= 25, filename = "Figures/varadero.skin.carotenoidvsyeast.png")
```
______
Sauce: orange carotenoid skin vs yeast orange skin
______
```{r}
sauce.skincount <- dplyr::select(countdata, contains("sauce"))
sauce.skinmorph <- deseqsamples_color %>% filter(morph == "sauce")

liverdds_all_sauce <- DESeqDataSetFromMatrix (countData = sauce.skincount,
                                       colData = sauce.skinmorph,
                                       design = ~ tissue + diet:tissue + tissue:pair.nn)

#run Wald test
liverdds.sauce.yellowVSblack <- DESeq(liverdds_all_sauce)
resultsNames(liverdds.sauce.yellowVSblack)

#extract the DEG between orange and black skin patches in the Sauce morph
res.sauce.yellowVSblack <- results(liverdds.sauce.yellowVSblack, name = c("tissueorange.dietyeast"))
res.sauce.yellowVSblack$transcript <- mcols(res.sauce.yellowVSblack)$transcript

# how many are "significant"?
table(res.sauce.yellowVSblack[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(res.sauce.yellowVSblack, 0.05, "sauce.orangevsorange_alpha_0.05")

```

______
Varadero: orange carotenoid skin vs yeast orange skin
______
```{r}
varadero.skincount <- dplyr::select(countdata, contains("varadero"))
varadero.skinmorph <- deseqsamples_color %>% filter(morph == "varadero")

dds_red_skin_varadero <- DESeqDataSetFromMatrix (countData = varadero.skincount,
                                       colData = varadero.skinmorph,
                                       design = ~ tissue + diet:tissue + tissue:pair.nn)

#run Wald test
dds.varadero.redVSred <- DESeq(dds_red_skin_varadero)
resultsNames(dds.varadero.redVSred)

#extract the DEG between orange and black skin patches in the Sauce morph
res.varadero.redVSred <- results(dds.varadero.redVSred, name = c("tissueorange.dietyeast"))
res.varadero.redVSred$transcript <- mcols(res.varadero.redVSred)$transcript

# how many are "significant"?
table(res.varadero.redVSred[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(res.varadero.redVSred, 0.05, "varadero.redvsred_alpha_0.05")

varadero.orange.skincount <- dplyr::select(countdata, contains("varadero_orange"))
varadero.orange.skinmorph <- deseqsamples_color %>% filter(tissue == "orange")
```

-------------
Yeast: orange vs red
-------------
```{r}
# create DESeq dataset for black vs color
dds.yeast.orangevsred <- DESeqDataSetFromMatrix (countData = countdata,
                                       colData = deseqsamples_color, 
                                       design = ~ morph + diet + tissue:diet)

#run Wald test
dds.yeast.orangevsred <- DESeq(dds.yeast.orangevsred, test="Wald")
resultsNames(dds.yeast.orangevsred)

#extract the DEG between black and orange/red skin patches in the both the Sauce and Varadero morph
dds.yeast.orangevsred <- results(dds.yeast.orangevsred, name =("dietyeast.tissueorange"))
dds.yeast.orangevsred$transcript <- mcols(dds.yeast.orangevsred)$transcript

# how many are "significant"?
table(dds.yeast.orangevsred[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(dds.yeast.orangevsred, 0.05, "yeast.orangevsred_alpha_0.05")
```

Carotenoid: orange vs red
```{r}
# create DESeq dataset for black vs color
dds.carotenoid.orangevsred <- DESeqDataSetFromMatrix (countData = countdata,
                                       colData = deseqsamples_color, 
                                       design = ~ morph + diet + tissue:diet)

#run Wald test
dds.carotenoid.orangevsred <- DESeq(dds.carotenoid.orangevsred, test="Wald")
resultsNames(dds.carotenoid.orangevsred)

#extract the DEG between black and orange/red skin patches in the both the Sauce and Varadero morph
dds.carotenoid.orangevsred <- results(dds.carotenoid.orangevsred, name =("dietcarotenoid.tissueorange"))
dds.carotenoid.orangevsred$transcript <- mcols(dds.carotenoid.orangevsred)$transcript

# how many are "significant"?
table(dds.carotenoid.orangevsred[,"padj"] < 0.05)

#color genes 
SigGeneWrapper(dds.carotenoid.orangevsred, 0.05, "carotenoid.orangevsred_alpha_0.05")
```

